Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-10-06 article.tex
Removed paragraphs describing GStreamer elements
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1377 ccdd05d5-56c1-461a-bf98-85f6876523b4

The \lloid\ algorithm is amenable to latency free, real-time implementation.  However, such a real-time search pipeline would require integration directly into the data acquisition and storage systems of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing low-latency, but not real-time, signal processing software in order to implement
the \lloid\ algorithm.

We have implemented a prototype of the low-latency filtering stage using an
open-source signal processing environment called
\gstreamer\footnote{\url{http://gstreamer.net/}}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of
\GW\ detector data to audio data it is not surprising that
\gstreamer\ is useful for our purpose. \gstreamer\ also provides some useful
stock signal processing elements such as resamplers and filters.  We have
extended the \gstreamer\ framework by developing a library called
\gstlal\footnote{\url{https://www.lsc-group.phys.uwm.edu/daswg/projects/gstlal.html}}
that provides elements for \GW\ data analysis.

\gstreamer\ pipelines typically operate with very low (in some consumer
applications, imperceptibly low) latency rather than in true real-time because
signals are partitioned into blocks of samples, or \emph{buffers}.  This affords
a number of advantages, including amortizing the overhead of passing signals
between elements and grouping together sequences of similar operations.
However, buffering a signal incurs a latency of up to one buffer length.  This
latency can be made small at the cost of some additional overhead by making the
buffers sufficiently small.

Our prototype pipeline, tested with GStreamer version 0.10.33, consists of the
following stages.

\paragraph{Decimation}

The whitened detector data is reduced to successively lower sample rates by
decimation. Decimation involves applying an anti-aliasing filter to the data,
and then down-sampling by deleting samples.  We use GStreamer's stock
\texttt{audioresample} element, which provides a FIR decimation filter with
a Kaiser-windowed sinc function kernel.

The detector data is provided at every power-of-two sample rate
required by the time slices described in \eqref{eq:time-slices}.  At this point,
the \gstreamer\ pipeline splits into several parallel branches to handle each
time slice.

\paragraph{Basis filters}

Next, these decimated data streams are fed into parallel \fir\ filter banks.
The basis filters are implemented using a \gstlal\ element called
\texttt{lal\_firbank}, which provides a single input, multiple output bank of
\fir\ filters from a matrix of filter coefficients.  Several instances of
\texttt{lal\_firbank} execute in parallel to handle each of the time slices.
Filter coefficients are taken from the nonzero samples of the time slices in
equation~\eqref{eq:time-sliced-templates}, with the zero-padding effectively
accomplished by appropriate queues, or delay lines.

\paragraph{Reconstruction}

From the outputs of the \SVD\ basis filters, we form the partial \SNR\ streams
for each time  slice by multiplying by the reconstruction matrices.  This is
accomplished with the \texttt{gstlal} element \texttt{lal\_matrixmixer}.

\paragraph{Interpolation}

In order to form the early-warning output from each time slice, we have to add
the partial \SNR\ to the early-warning output from the subsequent time slices.
If the next time slice does not have the same sample rate, its output must first
be interpolated.  The \gstreamer\ element \texttt{audioresample} is used again here.

\paragraph{\SNR\ accumulation}

The early-warning output for each time slice is formed by accumulating
interpolated early-warning output from the subsequent time slice.  This process
continues until we have worked our way to the \SNR\ of the original templates
at the full sample rate $f^0$.  In this way, the \lloid\ algorithm and this
implementation naturally leads to a simple early-warning pipeline.

@@

Next, keeping the \SVD\ tolerance fixed at $\left(1-10^{-6}\right)$ and the
length of the decimation filter fixed at $N^\shortdownarrow = 192$, we studied
the impact of the length $N^\shortuparrow$ of the interpolation filter on
mismatch.  The mismatch as a function of $N^\shortuparrow$ is shown in
Figure~\ref{fig:bw}b.  The mismatch saturates at $\sim$$2 \times 10^{-4}$ with
$N^\shortuparrow = 64$.  We find that a filter length of 16 is sufficient to
meet our target mismatch of 0.003.

>>>>

The \lloid\ algorithm is amenable to latency free, real-time implementation.  However, a real-time search pipeline would require integration directly into the data acquisition and storage systems of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing low-latency, but not real-time, signal processing software in order to implement
the \lloid\ algorithm.

We have implemented a prototype of the low-latency filtering stage using an
open-source signal processing environment called
\gstreamer\footnote{\url{http://gstreamer.net/}} (version 0.10.33).
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of
\GW\ detector data to audio data it is not surprising that
\gstreamer\ is useful for our purpose. \gstreamer\ also provides some useful
stock signal processing elements such as resamplers and filters.  We have
extended the \gstreamer\ framework by developing a library called
\gstlal\footnote{\url{https://www.lsc-group.phys.uwm.edu/daswg/projects/gstlal.html}}
that provides elements for \GW\ data analysis.

\gstreamer\ pipelines typically operate with very low (in some consumer
applications, imperceptibly low) latency rather than in true real-time because
signals are partitioned into blocks of samples, or \emph{buffers}.  This affords
a number of advantages, including amortizing the overhead of passing signals
between elements and grouping together sequences of similar operations.
However, buffering a signal incurs a latency of up to one buffer length.  This
latency can be made small at the cost of some additional overhead by making the
buffers sufficiently small.  In any case, buffering is a reasonable strategy for low latency \LIGO\ data analysis because, as we previously remarked, the \LIGO\ data acquisition system has a granularity of $1/16$~s.

@@

Next, keeping the \SVD\ tolerance fixed at $\left(1-10^{-6}\right)$ and the
length of the decimation filter fixed at $N^\shortdownarrow = 192$, we studied
the impact of the length $N^\shortuparrow$ of the interpolation filter on
mismatch.  We use GStreamer's stock
\texttt{audioresample} element, which provides a FIR decimation filter with
a Kaiser-windowed sinc function kernel.  The mismatch as a function of $N^\shortuparrow$ is shown in
Figure~\ref{fig:bw}b.  The mismatch saturates at $\sim$$2 \times 10^{-4}$ with
$N^\shortuparrow = 64$.  We find that a filter length of 16 is sufficient to
meet our target mismatch of 0.003.
